---
title: "RSV_250123"
output: html_document
date: "2025-01-19"
---

### load in packages
```{r}
library(arrow)
library(dplyr)
library(survival)
library(survminer)
```

### read in parquet database
```{r}
NY_pq <- open_dataset("R:/ParquetData/NY", format = "parquet")
#NY_pq_df <- as.data.frame(NY_pq)
```

### [Backup for SEDD]
NY_pq_SEDD <- open_dataset("/Volumes/home/HCUPdata-CC0941-MEDSPH/ParquetData/NY_SEDD", format = "parquet")
column_names <- NY_pq_SEDD$schema$names
print(column_names)

### Variables I need
- Birth Date
  [153] birth_year = "Birth year" ***
  [152] birth_month = "Birth month"                                                               
  [1] age_year = "Age in years at admission"                                                      
  [2] age_days = "Age in days (when age < 1 year)"                                                
  [3] age_months = "Age in months (when age < 11 years)"    
  
- Basic
  [47] Sex = "Indicator of sex"  
  [140] Race = "Race (uniform)"   
  [50] Ethnicity = "Hispanic ethnicity (uniform)"
  [104] Income_quartile = "Median household income state quartile for patient ZIP Code"
  [12] Death =  'Died during hospitalization'
  
- Admission and Discharge
  [5] Amonth='Admission month'
  [151] Ayear="Admission year"  
  [11] Dhour = "Discharge Hour"
  [16] Dmonth = "Discharge month"
  [17] Dquarter = "Discharge quarter"
  [98] Length_Of_Stay_C = "Length of stay (cleaned)"
  [99] Length_Of_Stay_O = "Length of stay (as received from source)"
 
- Geographic
  [148] ZIP="Patient ZIP Code"
  [149] ZIP_3digits = "Patient ZIP Code, frist 3 digits"
  [112] Loc_CBSA = "Patient location: Core Based Statistical Area (CBSA)"                            
  [113] Loc_UrbanRural = "Patient Location: NCHS Urban-Rural Code"                                   
  [114] Loc_UrbanInfluence = "Patient location: Urban Influence Codes"                            
  [115] Loc_UrbanRural4Cate = "Patient Location: Urban-Rural 4 Categories"  
  [138] State = "Patient State postal code"                                                          
  [139] State_FIPS = "Patient state/county FIPS code" 
  
- Identifier
  [146] ID="Visit linkage variable"
  
- Variables
  [9] BWT = "Birth weight in grams"
  [150] Income_Quartile = "Median household income national quartile for patient ZIP Code"                           
- Stay Service
  [6] AdminType = "Admission type"
  [49] StayService = "HCUP Observation Stay service indicator"
  [144] TransIn = "Transfer in indicator"
  [145] TransOut = "Transfer out indicator"
  
- Diagnosis
  [22-46] Diagnosis_N = "Diagnosis N, present on admission indicator"  
  [54-96] ICD_N = "ICD Code"
 
 
### [Mute] Compute the range for the "Birth year" column
 col_range <- NY_pq %>%
  select(`Birth year`) %>%  
  collect() %>%            
  pull() %>%              
  range(na.rm = TRUE) 
  
### [Mute] Display the range
col_range # Result: 1894 2019, 


### [Mute] Cohort 1 include every variables I need, birthday already fixed from 2014-2020

 NY_cohort1 <- NY_pq_df %>%
  rename(
# Basic and birthday
    ID = 'Visit linkage variable',
    Amonth = 'Admission month',
    Ayear = 'Admission year',
    Dhour = 'Discharge Hour',
    Dmonth = 'Discharge month',
    Dquarter = 'Discharge quarter',
    Dyear = 'Calendar year',
    Length_Of_Stay_C = 'Length of stay (cleaned)',
    Length_Of_Stay_O = 'Length of stay (as received from source)',
    birth_year = 'Birth year',
    birth_month = 'Birth month',
    admis_age_year = 'Age in years at admission',
    admis_age_day = 'Age in days (when age < 1 year)',
    admis_age_mon = 'Age in months (when age < 11 years)',
    Sex = 'Indicator of sex',
    Race = 'Race (uniform)',
    Ethnicity = 'Hispanic ethnicity (uniform)',
    Income_state_quartile = 'Median household income state quartile for patient ZIP Code',
    Income_national_quartile = 'Median household income national quartile for patient ZIP Code',
    BWT = 'Birth weight in grams',
# Location
    ZIP = 'Patient ZIP Code',
    ZIP_3digits = 'Patient ZIP Code, first 3 digits',
    Loc_CBSA = 'Patient location: Core Based Statistical Area (CBSA)',
    Loc_UrbanRural = 'Patient Location: NCHS Urban-Rural Code',
    Loc_UrbanInfluence = 'Patient location: Urban Influence Codes',
    Loc_UrbanRural4Cate = 'Patient Location: Urban-Rural 4 Categories',
    State = 'Patient State postal code',
    State_FIPS = 'Patient state/county FIPS code',
# Service and diagnosis
    AdminType = "Admission type",
    StayService = 'HCUP Observation Stay service indicator',
    TransIn = 'Transfer in indicator',
    TransOut = 'Transfer out indicator',
    Death = 'Died during hospitalization',
    Diagnosis_1 = "Diagnosis 1, present on admission indicator",
    Diagnosis_2 = "Diagnosis 2, present on admission indicator",
    Diagnosis_3 = "Diagnosis 3, present on admission indicator",
    Diagnosis_4 = "Diagnosis 4, present on admission indicator",
    Diagnosis_5 = "Diagnosis 5, present on admission indicator",
    Diagnosis_6 = "Diagnosis 6, present on admission indicator",
    Diagnosis_7 = "Diagnosis 7, present on admission indicator",
    Diagnosis_8 = "Diagnosis 8, present on admission indicator",
    Diagnosis_9 = "Diagnosis 9, present on admission indicator",
    Diagnosis_10 = "Diagnosis 10, present on admission indicator",
    Diagnosis_11 = "Diagnosis 11, present on admission indicator",
    Diagnosis_12 = "Diagnosis 12, present on admission indicator",
    Diagnosis_13 = "Diagnosis 13, present on admission indicator",
    Diagnosis_14 = "Diagnosis 14, present on admission indicator",
    Diagnosis_15 = "Diagnosis 15, present on admission indicator",
    Diagnosis_16 = "Diagnosis 16, present on admission indicator",
    Diagnosis_17 = "Diagnosis 17, present on admission indicator",
    Diagnosis_18 = "Diagnosis 18, present on admission indicator",
    Diagnosis_19 = "Diagnosis 19, present on admission indicator",
    Diagnosis_20 = "Diagnosis 20, present on admission indicator",
    Diagnosis_21 = "Diagnosis 21, present on admission indicator",
    Diagnosis_22 = "Diagnosis 22, present on admission indicator",
    Diagnosis_23 = "Diagnosis 23, present on admission indicator",
    Diagnosis_24 = "Diagnosis 24, present on admission indicator",
    Diagnosis_25 = "Diagnosis 25, present on admission indicator",
    ICD_Admin = 'ICD-10-CM Admitting Diagnosis Code',
    ICD_D1 = 'ICD-10-CM Diagnosis 1',
    ICD_D2 = 'ICD-10-CM Diagnosis 2',
    ICD_D3 = 'ICD-10-CM Diagnosis 3',
    ICD_D4 = 'ICD-10-CM Diagnosis 4',
    ICD_D5 = 'ICD-10-CM Diagnosis 5',
    ICD_D6 = 'ICD-10-CM Diagnosis 6',
    ICD_D7 = 'ICD-10-CM Diagnosis 7',
    ICD_D8 = 'ICD-10-CM Diagnosis 8',
    ICD_D9 = 'ICD-10-CM Diagnosis 9',
    ICD_D10 = 'ICD-10-CM Diagnosis 10',
    ICD_D11 = 'ICD-10-CM Diagnosis 11',
    ICD_D12 = 'ICD-10-CM Diagnosis 12',
    ICD_D13 = 'ICD-10-CM Diagnosis 13',
    ICD_D14 = 'ICD-10-CM Diagnosis 14',
    ICD_D15 = 'ICD-10-CM Diagnosis 15',
    ICD_D16 = 'ICD-10-CM Diagnosis 16',
    ICD_D17 = 'ICD-10-CM Diagnosis 17',
    ICD_D18 = 'ICD-10-CM Diagnosis 18',
    ICD_D19 = 'ICD-10-CM Diagnosis 19',
    ICD_D20 = 'ICD-10-CM Diagnosis 20',
    ICD_D21 = 'ICD-10-CM Diagnosis 21',
    ICD_D22 = 'ICD-10-CM Diagnosis 22',
    ICD_D23 = 'ICD-10-CM Diagnosis 23',
    ICD_D24 = 'ICD-10-CM Diagnosis 24',
    ICD_D25 = 'ICD-10-CM Diagnosis 25',
    ICD_P1 = 'ICD-10-PCS Procedure 1',
    ICD_D_num = 'ICD-10-CM Number of diagnoses on this record',
    ICD_P_num = 'ICD-10-PCS Number of procedures on this record',
    ICD_P2 = 'ICD-10-PCS Procedure 2',
    ICD_P3 = 'ICD-10-PCS Procedure 3',
    ICD_P4 = 'ICD-10-PCS Procedure 4',
    ICD_P5 = 'ICD-10-PCS Procedure 5',
    ICD_P6 = 'ICD-10-PCS Procedure 6',
    ICD_P7 = 'ICD-10-PCS Procedure 7',
    ICD_P8 = 'ICD-10-PCS Procedure 8',
    ICD_P9 = 'ICD-10-PCS Procedure 9',
    ICD_P10 = 'ICD-10-PCS Procedure 10',
    ICD_P11 = 'ICD-10-PCS Procedure 11',
    ICD_P12 = 'ICD-10-PCS Procedure 12',
    ICD_P13 = 'ICD-10-PCS Procedure 13',
    ICD_P14 = 'ICD-10-PCS Procedure 14',
    ICD_P_type = 'ICD-10-PCS Procedure type indicator'
  ) %>%
  filter(!is.na(ID), BWT != '', birth_year >= 2014, birth_year <= 2020) %>%
  select(
    ID, birth_year, birth_month, Ayear, Amonth, admis_age_year, admis_age_mon, 
    admis_age_day, Dhour, Dmonth, Dquarter, Dyear, Length_Of_Stay_C, Length_Of_Stay_O,
    Sex, Race, Ethnicity, Income_state_quartile, Income_national_quartile, BWT, 
    State, State_FIPS, ZIP, ZIP_3digits, Loc_CBSA, Loc_UrbanRural, 
    Loc_UrbanInfluence, Loc_UrbanRural4Cate,    
    AdminType, StayService, TransIn, TransOut, Death, 
    Diagnosis_1, Diagnosis_2, Diagnosis_3, Diagnosis_4, Diagnosis_5, 
    Diagnosis_6, Diagnosis_7, Diagnosis_8, Diagnosis_9, Diagnosis_10, 
    Diagnosis_11, Diagnosis_12, Diagnosis_13, Diagnosis_14, Diagnosis_15, 
    Diagnosis_16, Diagnosis_17, Diagnosis_18, Diagnosis_19, Diagnosis_20, 
    Diagnosis_21, Diagnosis_22, Diagnosis_23, Diagnosis_24, Diagnosis_25, 
    ICD_D1, ICD_D2, ICD_D3, ICD_D4, ICD_D5, 
    ICD_D6, ICD_D7, ICD_D8, ICD_D9, ICD_D10, 
    ICD_D11, ICD_D12, ICD_D13, ICD_D14, ICD_D15, 
    ICD_D16, ICD_D17, ICD_D18, ICD_D19, ICD_D20, 
    ICD_D21, ICD_D22, ICD_D23, ICD_D24, ICD_D25, 
    ICD_P1, ICD_P2, ICD_P3, ICD_P4, ICD_P5, 
    ICD_P6, ICD_P7, ICD_P8, ICD_P9, ICD_P10, 
    ICD_P11, ICD_P12, ICD_P13, ICD_P14, 
    ICD_Admin, ICD_D_num, ICD_P_num, ICD_P_type
  ) %>%
  collect()

# Save the dataset to an RDS file
saveRDS(NY_cohort1, '/Volumes/home/HCUPdata-CC0941-MEDSPH/myrrha/Data/total_cohort.rds')


### Skip all former steps, direct write in total cohort as NY_cohort1
```{r}
NY_cohort1 <- readRDS("R:/myrrha/Data/total_cohort.rds")
```

### Cohort 2 add on: age_norm ; rsv_indi and details using ICD code
```{r}

# Turn age to numeric and normalize to day unit

NY_cohort2 <- NY_cohort1 %>%
  # Filter out rows with non-numeric values
  filter(
    birth_year>=2016, #there are a handful of births in 2014/15 that should be excluded
    !admis_age_year %in% c( "A", "B", "C"),
    !admis_age_mon %in% c( "A", "B", "C"),
    !admis_age_day %in% c( "A", "B", "C")
  ) %>%
  # Convert columns to numeric after filtering
  mutate(
    admis_age_year = as.numeric(admis_age_year),
    admis_age_mon = as.numeric(admis_age_mon),
    admis_age_day = as.numeric(admis_age_day),
  )

NY_cohort3 <- NY_cohort2 %>%
  filter(
    admis_age_year < 2 | is.na(admis_age_year),
    admis_age_mon < 24 | is.na(admis_age_mon),
    admis_age_day < 730 | is.na(admis_age_day)
  ) %>%
  filter(
    !is.na(admis_age_year) | !is.na(admis_age_mon) | !is.na(admis_age_day)
  ) %>%
  mutate(
    admis_age_norm = pmax(
      admis_age_year * 365.25,
      admis_age_mon * 30.44,
      admis_age_day,
      na.rm = TRUE
    )
  )

# RSV indicator as a new column
NY_cohort3 <- NY_cohort3 %>%
  mutate(
    rsv_bronchitis = if_any(`ICD_D1`:`ICD_D25`, ~ . == "J205"),
    rsv_bronchiolitis = if_any(`ICD_D1`:`ICD_D25`, ~ . == "J210"),
    rsv_pneumonia = if_any(`ICD_D1`:`ICD_D25`, ~ . == "J121"),
    rsv_cause_of_disease = if_any(`ICD_D1`:`ICD_D25`, ~ . == "B974")
  ) %>%
  mutate(across(starts_with("rsv_"), as.integer)) %>% # Convert logical to 1/0
  mutate(
    rsv_indi = if_else(
      rsv_bronchitis == 1 | rsv_bronchiolitis == 1 | rsv_pneumonia == 1 | rsv_cause_of_disease == 1, 1,
      0
    )
  )

```

### [Mute] Discharge_check
Discharge_check <- NY_cohort3 %>%
  select(ID, birth_year, birth_month, 
  admis_age_year, admis_age_mon, admis_age_day, admis_age_norm, 
  Length_Of_Stay_C, Length_Of_Stay_O)

### Finds rsv encounters each patient has and flag as 1, count number
```{r}
NY_cohort4 <- NY_cohort3 %>%
  mutate(across(
    ICD_D1:ICD_D25,
    ~ ifelse(. %in% c("J205", "J210", "J121", "B974"), 1, 0),
    .names = "rsv_{.col}"
  )) %>%
  select(everything(), starts_with("rsv_ICD"), rsv_indi)

# Total encounter of RSV
NY_cohort4 <- NY_cohort4 %>%
  mutate(rsv_num = rowSums(select(., starts_with("rsv_ICD"))))
```

### Flag rows of patient who were admitted with rsv
```{r}
NY_cohort4 <- NY_cohort4 %>%
  mutate(across(starts_with("Diagnosis_"), ~ as.integer(. == "Y"), .names = "check_{.col}"))

NY_cohort4 <- NY_cohort4 %>%
  rowwise() %>%
  mutate(
    rsv_admis = if_else(
      any(c_across(rsv_ICD_D1:rsv_ICD_D25) == 1 & c_across(check_Diagnosis_1:check_Diagnosis_25) == 1),
      1,
      0
    )
  ) %>%
  ungroup()

saveRDS(NY_cohort4, '/Volumes/home/HCUPdata-CC0941-MEDSPH/myrrha/Data/NY_cohort4.rds')
```

### Just to make sure the rsv_admis is coded correct
rsv_check <- NY_cohort4 %>%
  select(
    ID, rsv_admis,
    rsv_ICD_D1, check_Diagnosis_1,
    rsv_ICD_D2, check_Diagnosis_2,
    rsv_ICD_D3, check_Diagnosis_3,
    rsv_ICD_D4, check_Diagnosis_4,
    rsv_ICD_D5, check_Diagnosis_5,
    rsv_ICD_D6, check_Diagnosis_6,
    rsv_ICD_D7, check_Diagnosis_7,
    rsv_ICD_D8, check_Diagnosis_8,
    rsv_ICD_D9, check_Diagnosis_9,
    rsv_ICD_D10, check_Diagnosis_10,
    rsv_ICD_D11, check_Diagnosis_11,
    rsv_ICD_D12, check_Diagnosis_12,
    rsv_ICD_D13, check_Diagnosis_13,
    rsv_ICD_D14, check_Diagnosis_14,
    rsv_ICD_D15, check_Diagnosis_15,
    rsv_ICD_D16, check_Diagnosis_16,
    rsv_ICD_D17, check_Diagnosis_17,
    rsv_ICD_D18, check_Diagnosis_18,
    rsv_ICD_D19, check_Diagnosis_19,
    rsv_ICD_D20, check_Diagnosis_20,
    rsv_ICD_D21, check_Diagnosis_21,
    rsv_ICD_D22, check_Diagnosis_22,
    rsv_ICD_D23, check_Diagnosis_23,
    rsv_ICD_D24, check_Diagnosis_24,
    rsv_ICD_D25, check_Diagnosis_25
  )

###
```{r}

all_birth <- NY_cohort4 %>%
  select(ID, birth_year, birth_month)

rsv_diagnosis_vari <- NY_cohort4 %>%
  select(ID, Ayear, Amonth, Dyear, Dmonth, admis_age_norm, rsv_admis, rsv_indi, Sex, Race, Ethnicity, Income_state_quartile, Income_national_quartile, BWT,ZIP)


saveRDS(all_birth, '/Volumes/home/HCUPdata-CC0941-MEDSPH/myrrha/Data/all_birth.rds')
saveRDS(rsv_diagnosis_vari, '/Volumes/home/HCUPdata-CC0941-MEDSPH/myrrha/Data/rsv_diagnosis_vari.rds')
```

### Whole KM curve
```{r}
km_data <- merge(all_birth, rsv_diagnosis_vari, by = "ID")

km_data <- km_data %>%
  mutate(
    go_to_event = if_else(
      rsv_admis == 1,
      (Ayear - birth_year) * 12 + (Amonth - birth_month),
      (Dyear - birth_year) * 12 + (Dmonth - birth_month)
    )
  )

# Create a survival object
surv_obj <- Surv(time = km_data$go_to_event, event = km_data$rsv_admis)

# Fit the KM model
km_fit <- survfit(surv_obj ~ 1, data = km_data)

# Plot the KM curve
ggsurvplot(km_fit, km_data = km_data, conf.int = TRUE, risk.table = TRUE,
           xlab = "Time (Months)", ylab = "Survival Probability")

```

# Variables clean up
```{r}
km_data <- km_data %>%
  mutate(
    Sex = case_when(
      Sex == "0" ~ "Male",
      Sex == "1" ~ "Female",
      TRUE ~ NA_character_
    ),
    Race = case_when(
      Race == "1" ~ "White",
      Race == "2" ~ "Black",
      Race == "3" ~ "Hispanic",
      Race == "4" ~ "Asian or Pacific Islander",
      Race == "5" ~ "Native American",
      Race == "6" ~ "Other",
      TRUE ~ NA_character_
    ),
    Ethnicity = case_when(
      Ethnicity == 0 ~ "Not Hispanic",
      Ethnicity == 1 ~ "Hispanic, White",
      Ethnicity == 2 ~ "Hispanic, Black",
      Ethnicity == 3 ~ "Hispanic, Other Race",
      Ethnicity == 4 ~ "Hispanic, Unspecified Race",
      TRUE ~ NA_character_ 
    ),
    N_Quartile = case_when(
      Income_national_quartile == 1 ~ "First Quartile",
      Income_national_quartile == 2 ~ "Second Quartile",
      Income_national_quartile == 3 ~ "Third Quartile",
      Income_national_quartile == 4 ~ "Fourth Quartile",
      TRUE ~ NA_character_ 
    ),
    S_Quartile = case_when(
      Income_state_quartile == 1 ~ "First Quartile",
      Income_state_quartile == 2 ~ "Second Quartile",
      Income_state_quartile == 3 ~ "Third Quartile",
      Income_state_quartile == 4 ~ "Fourth Quartile",
      TRUE ~ NA_character_ 
    )
  )
```

```{r}
km_data$Sex <- as.factor(km_data$Sex)
km_data$Race <- as.factor(km_data$Race)
km_data$Ethnicity <- as.factor(km_data$Ethnicity)
km_data$N_Quartile <- as.factor(km_data$N_Quartile)
km_data$S_Quartile <- as.factor(km_data$S_Quartile)

```

```{r}
# Fit the KM model stratified by Sex
km_fit_sex <- survfit(surv_obj ~ Sex, data = km_data)

# Plot the KM curve stratified by Sex
ggsurvplot(
  km_fit_sex,
  data = km_data,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Curve Stratified by Sex"
)
```
```{r}
# Fit the KM model stratified by Race
km_fit_race <- survfit(surv_obj ~ Race, data = km_data)

# Plot the KM curve stratified by Race
ggsurvplot(
  km_fit_race,
  data = km_data,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Curve Stratified by Race"
)
```
```{r}
# Fit the KM model stratified by Race
km_fit_ethnicity <- survfit(surv_obj ~ Ethnicity, data = km_data)

# Plot the KM curve stratified by Race
ggsurvplot(
  km_fit_ethnicity,
  data = km_data,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Curve Stratified by Ethnicity"
)
```

```{r}
# Fit the KM model stratified by Race
km_fit_n_income_Quartile <- survfit(surv_obj ~ N_Quartile, data = km_data)

# Plot the KM curve stratified by Race
ggsurvplot(
  km_fit_n_income_Quartile,
  data = km_data,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Curve Stratified by National Income Quartile"
)
```
```{r}
# Fit the KM model stratified by Race
km_fit_s_income_Quartile <- survfit(surv_obj ~ S_Quartile, data = km_data)

# Plot the KM curve stratified by Race
ggsurvplot(
  km_fit_s_income_Quartile,
  data = km_data,
  conf.int = TRUE,
  risk.table = TRUE,
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  title = "Kaplan-Meier Curve Stratified by State Income Quartile"
)
```
